\documentclass[en]{pracamgr}

\usepackage{float}
\usepackage{xpatch}
\usepackage{listings}
\usepackage{minted}
\usepackage{tikz-cd}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{epigraph}
\usepackage{amsfonts}
\usepackage{ragged2e}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{breakcites}

\input{customs.tex}

\autor{Radosław Jan Rowicki}{386088}

\title{Liquid types for verification of smart contracts}
\titlepl{Weryfikacja inteligentnych kontraktów przy pomocy logicznie
  kwalifikowanych typów danych}
\kierunek{Computer Science}

\opiekun{PhD. Ulf Norell\\
  University of Gothenburg\\
}
\opiekunuw{PhD. Marcin Benke\\
  University of Warsaw\\
}

\date{September 2021}

%Podać dziedzinę wg klasyfikacji Socrates-Erasmus:
\dziedzina{
  11.3 Informatics
}

\klasyfikacja{
  D Software\\
  D.1 Programming Techniques\\
  D.1.1 Functional Programming\\
  D.3 Programming Languages\\
  D.3.3 Language Constructs and Features\\
  F Theory of Computation\\
  F.3 Logics and Meanings of Programs\\
  F.3.2 Specifying and Verifying and Reasoning about Programs }

\keywords{liquid types, smart contracts, formal verification, dependent types,
  functional programming, blockchain}

\newtheorem{defi}{Definition}[section]

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

\begin{document}

\maketitle

\center \subsection*{Acknowledgements}

\justify First of all, I would like to thank my supervisor, Ulf Norell, for
supporting me during my work by inspiring me with ideas, giving numerous
valuable comments, and discussing various solutions. I am also grateful for our
cooperation in developing the compiler of Sophia, during which I learned a lot
about implementing programming languages. Next, I appreciate the support from
Marcin Benke representing the University of Warsaw, who coordinated the formal
side of my thesis. Furthermore, I am thankful for what he taught me about type
systems and semantics of functional languages.

\justify This work has been a subject of the 0135 grant of the Aeternity Crypto
Foundation ``Logic Qualified Data Types for Sophia''. I hereby thank Lydia
Atanassova representing the Foundation for her assistance regarding the grant
application. I would also like to thank all my colleagues from the {\ae}ternity
Core--Dev team for being exceptionally inspiring professionals, with whom it was
a big pleasure to work.

\justify Last but not least, I would like to thank my girlfriend, Karolina
Tudelska, for her moral support and for providing me with priceless advice on
writing texts in English.

\begin{abstract}
  In recent years smart contracts have been developed within the blockchain
  technology as a trustless tool for managing digital assets, many of which are
  of considerable value. Due to the nature of blockchains, smart contracts are
  vulnerable to bugs, because once deployed, they cannot be fixed. This can have
  wide-ranging financial consequences, as cryptocurrencies and NFTs rapidly gain
  popularity. Therefore, a great need for meticulous audits emerges, so the
  vulnerabilities can be found and mitigated before they cause harm.

  This work presents liquid types as a method of computer-aided verification of
  smart contracts. The analysed implementation named Hagia is an extension of
  the type system of the Sophia language, which has been designed for smart
  contract development on the {\ae}ternity blockchain. Liquid types have already
  found application in many projects written in functional languages such as
  OCaml or Haskell. The presented algorithms and discussions provide a framework
  for understanding the key concepts of the liquid inference, use cases and the
  benefits of the proposed solution in the context of smart contract
  development.
\end{abstract}

\input{changelog.tex}

\tableofcontents

\input{chapters/introduction.tex}
\input{chapters/sophia.tex}
\input{chapters/liquid_types.tex}
\input{chapters/implementation.tex}
\input{chapters/outcome.tex}
\input{chapters/discussion.tex}
\input{chapters/conclusions.tex}

\appendix
\input{chapters/appendix.tex}

\listoffigures

\bibliographystyle{acm} \bibliography{bibliography}
\end{document}
